---
title: "Playing with domestic airline perfomance data"
author: "Kirill"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    df_print: paged
    code_folding: "hide"
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: sandstone
    highlight: espresso
---

```{r setup, echo = F}
options(encoding="utf-8")
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

## Introduction

I found this rather interesting data set at [data.gov.au](https://data.gov.au), [Domestic Airlines - On Time Performance](https://data.gov.au/data/dataset/domestic-airline-on-time-performance) and I decided to investigate it a bit closer.

First thing first is to download it. Note that `read_csv` from [`readr`](https://readr.tidyverse.org/) package can "read" directly from url, but I wasn't sure if everytime I compile html it would re-download the file or use chached version. The data is _Creative Commons Attribution 3.0 Australia_ and so no problem in downloading and using the data.

### Loading libraries

We are going to use

```{r message = F}
library(tidyverse)
library(knitr)
```

### Downloading the data

We are going to use [`tidyverse`](https://www.tidyverse.org/) library that includes several other useful libraries, such as:

- `readr`
- `tidyr`
- `dplyr`
- `ggplot2`

to name a few

Note that we are doing conditional download here, obviously don't want to re-download if we already have the file.

```{r}
fn_data <- "domestic_airline_performance.csv"
fn_notes <- "domestic_airline_performance_notes.txt"

if(!file.exists(fn_data)) {
  url_data <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/cf663ed1-0c5e-497f-aea9-e74bfda9cf44/download/otp_time_series_web.csv"
  url_notes <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/69e214b9-b66f-4636-9c84-3efda2d0d8ae/download/otptimeserieswebnotes.txt"

  download.file(url_data, fn_data)
  download.file(url_notes, fn_notes)
}

df <- read_csv(fn_data, quote = "")

df
```

## Exploring the data

Now that we've got the data lets explore it. It always helps if we can find more information about the data set, particular what information each column might have.

### Working with data

Great, the information above gives us some starting material. However it wasn't that explicit what each column meant and how man columns are there. Let's quickly take a pick

```{r}
d <- df %>% dim
```

total number of observation `r d[1]` and total number of variables `r d[2]`

There are many ways you can explore this data, but i just want to have a look at the types of Airlines there are.

```{r}
df %>%
  select(Airline) %>%
  distinct() %>%
  arrange(Airline)
```

## Cleaning up

I've noticed that there "All Airlines" name in the `Airline` column that appears to have the most number of occurrences in the data

```{r}
df %>%
  group_by(Airline) %>%
  summarise(n = n()) %>%
  arrange(-n)
```

Also there one of the routes is `All Ports-All Ports`. Googling for that name didn't reveal any places in Australia by that name.

```{r}
df %>%
  group_by(Route) %>%
  summarise(n = n()) %>%
  arrange(-n)
```

I decide going forward to drop those data points.

```{r}
df2 <- df %>%
          filter(Airline != "All Airlines",
                 Route != "All Ports-All Ports")
```


## Visualising the data

Here we are summarising so that we have an idea of how many times a particular location had be use per airline per year and we are only going to look at two airlines,
Jetstar and Qantas.

```{r}
p2 <- df2 %>%
  group_by(Airline, Year, Departing_Port) %>%
  summarise(n = n()) %>%
  ungroup %>%
  filter(Airline == "Jetstar" | Airline == "Qantas") %>%
  ggplot(aes(Departing_Port, n, color = Airline)) +
    geom_boxplot() +
    theme(axis.text.x=element_text(angle=45, hjust=1))

p2
```

In any given year what is the distribution of cancellation

```{r, fig.width=15, fig.height=9}
p3 <- df2 %>%
        filter(Airline == "Jetstar" | Airline == "Qantas") %>%
        select(Airline, Departing_Port, Cancellations, Year) %>%
        ggplot(aes(Departing_Port, Cancellations, color = factor(Year))) +
          geom_boxplot() +
          facet_wrap(~Airline) +
          theme(axis.text.x=element_text(angle= 45, hjust=1))
p3
```

## References

- [themes](http://www.datadreaming.org/post/r-markdown-theme-gallery/)
