---
title: "Playing with domestic airline perfomance data"
author: "Kirill"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: sandstone
---

## Introduction

I found this rather interesting data set at [data.gov.au](https://data.gov.au), [Domestic Airlines - On Time Performance](https://data.gov.au/data/dataset/domestic-airline-on-time-performance) and I decided to investigate it a bit closer.

First thing first is to download it. Note that `read_csv` from [`readr`](https://readr.tidyverse.org/) package can "read" directly from url, but I wasn't sure if everytime I compile html it would re-download the file or use chached version. The data is _Creative Commons Attribution 3.0 Australia_ and so no problem in downloading and using the data.

### Loading libraries

We are going to use
```{r message = F}
library(tidyverse)
library(knitr)
```

### Downloading the data

We are going to use [`tidyverse`](https://www.tidyverse.org/) library that includes several other useful libraries, such as:

- `readr`
- `tidyr`
- `dplyr`
- `ggplot2`

to name a few

Note that we are doing conditional download here, obviously don't want to re-download if we already have the file.

```{r}
fn_data <- "domestic_airline_performance.csv"
fn_notes <- "domestic_airline_performance_notes.txt"

if(!file.exists(fn_data)) {
  url_data <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/cf663ed1-0c5e-497f-aea9-e74bfda9cf44/download/otp_time_series_web.csv"
  url_notes <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/69e214b9-b66f-4636-9c84-3efda2d0d8ae/download/otptimeserieswebnotes.txt"

  download.file(url_data, fn_data)
  download.file(url_notes, fn_notes)
}

df <- read_csv(fn_data, quote = "")

df

df %>% head %>% kable
```

`kable` is a helper function to help us display tables in Rmarkdown

## Exploring the data

Now that we've got the data lets explore it. It always helps if we can find more information about the data set, particular what information each column might have.

### External notes about the data

```{r child="domestic_airline_performance_notes.txt"}

```

***

### Working with data

Great, the information above gives us some starting material. However it wasn't that explicit what each column meant and how man columns are there. Let's quickly take a pick

```{r}
df %>% dim
```

The reason I'm going to converted to `data.frame` is simply becasue I want to see one entry per line

```{r warning=F}
df %>%
  colnames %>%
  unlist %>%
  sort %>%
  data.frame
```

There are many ways you can explore this data, but i just want to have a look at the types of Airlines there are.

```{r}
df %>%
  select(Airline) %>%
  distinct() %>%
  arrange(Airline)
```

## Cleaning up

I've noticed that there "All Airlines" name in the `Airline` column that appears to have the most number of occurrences in the data

```{r}
df %>%
  group_by(Airline) %>%
  summarise(n = n()) %>%
  arrange(-n)
```

Also there one of the routes is `All Ports-All Ports`. Googling for that name didn't reveal any places in Australia by that name.

```{r}
df %>%
  group_by(Route) %>%
  summarise(n = n()) %>%
  arrange(-n)
```

I decide going forward to drop those data points.

```{r}
df2 <- df %>%
          filter(Airline != "All Airlines",
                 Route != "All Ports-All Ports")
```


## Visualising the data

### Bar plot

```{r}
df2 %>%
  group_by(Airline, Route) %>%
  summarise(n = n()) %>%
  ungroup %>%
  ggplot(aes(Route, n)) +
    geom_bar(stat = 'identity') +
    facet_wrap(~Airline, ncol = 5, scales = "free_y") +
    theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Scatter plot

```{r}
p2 <- df2 %>%
  group_by(Airline, Year, Departing_Port, Arriving_Port) %>%
  summarise(n = n()) %>%
  ungroup %>%
  ggplot(aes(Departing_Port, Arriving_Port, color = factor(Year))) +
    geom_point(aes(size = n)) +
    facet_wrap(~Airline, ncol = 5, scales = "free") +
    theme(axis.text.x=element_text(angle=45, hjust=1))

p2
```

```{r}
p3 <- df2 %>%
        select(Airline, Route, Cancellations, Year, Month_Num) %>%
        ggplot(aes(factor(Year), Cancellations, color = Airline)) +
          geom_point() +
          facet_wrap(~Airline, ncol = 5) +
          theme(axis.text.x=element_text(angle= 45, hjust=1))
p3
```

```{r}

k <- 10

p4 <- df2 %>%
        select(Airline, Route, Cancellations, Year, Month_Num) %>%
        group_by(Airline, Year) %>%
        mutate(tot = n()) %>%
        ungroup %>%
        mutate(norm_cancel = (Cancellations/tot) * k) %>%
        ggplot(aes(factor(Year), norm_cancel, color = Airline)) +
          geom_point() +
          facet_wrap(~Airline, ncol = 5) +
          theme(axis.text.x=element_text(angle=45, hjust=1))
p4
```

It appears that after normalising for number of flights in that year by a particular Airline that Virgin Aus, ATR has the higest number of cancellations.

Need to check but I think the intepretation of the my normalisation is that for every 10 flights 6 are cancelled

## Next step

Learn from the data which airlines appear to be departing/arrivign late and to which destinations

## References

- [themes](http://www.datadreaming.org/post/r-markdown-theme-gallery/)
