# R code chunks

[Knitr R package](https://cran.r-project.org/web/packages/knitr/) is pivotal in providing mechanism for including **live** code into the document. Thus far we have learned how to include blobs of code into our documents and even do simple code highlighting by using language keywords at the end of the triple backticks. with `Knitr` however, we can also execute a.k.a run the code inside the document and in additional to the blob of code, also display the result from it. This is extremely useful for research communication and reproducibility. 

For this part of the book, we will need to install a few additional packages. Copy and paste install packages code into your RStudio console

```{r eval = FALSE}
install.packages(c("dplyr", "DT", "ggplot2"))
```

## Introduction

Each code block is referred to as "code chunk". You can have as many code chunks as you like. The code chunk is essentially what we have already been using, i.e triple backticks on either side of the code blob, but now add curly brackets at the end of the first triplet e.g

````markdown
`r ''````{r}

`r ''````
````

Look for green "insert" button at the top of the text editor pane, in order to insert R chunk. Alternatively use keyboard shortcut `alt+shift+i` 

Each chunk is highly configurable via chunk options, using key value pairs once again. The very first chunk option is `engine` i.e which language is included in the code chunk. Yes Rmarkdown is capable of doing many different languages. Use drop down arrow on the green "insert" button to explore some of the most common languages.

**N.B:** Name of the engine i.e the language name, always position first in the R chunk, usually key "engine" is omitted.

Let's include our first bit of R code. We are going to load a library `magrittr`, to get a pipe operator (it isn't overly important to understand the pipe operator for this course), then we are going to call base function `avaliable.packages()` and store results in the variable `avail_packages` and `Knit` our document once again. Note that I'm also including new section header.

````markdown
# R code

`r ''````{r}
library(magrittr)

avail_packages <- available.packages(contriburl = contrib.url("https://mirror.aarnet.edu.au/pub/CRAN/"))

#head(avail_packages)

avail_packages %>% head
`r ''````

````

--- It's computing !

## Chunk name 

It is very useful to name your code chunks, they can server multiple purposes such as cross-referencing and debugging. Names are optional, but if you are including them, they must be in a second position.

General layout of any chunk is

````markdown
`r ''````{r chunk_name, options}
`r ''````
````

Let's add a `get_data` name to our code chunk. We also going to wrangle our data a little bit to get most relevant information for us and press `Knit`

````markdown
`r ''````{r get_data}

library(magrittr)
library(dplyr)

avail_packages <- available.packages(contriburl = contrib.url("https://mirror.aarnet.edu.au/pub/CRAN/")) %>% 
                    as.data.frame() %>% 
                    as_tibble() %>% 
                    rowwise() %>% 
                    mutate(Depend = length(unlist(strsplit(Imports, split = ",")))) %>% 
                    ungroup() %>% 
                    select(Package, Version, Depend, Imports, License)
`r ''````
````

--- How is it looking?

## Chunk options

For comprehensive overview of code chunks that `knitr` refer to [definitive guide](https://yihui.name/knitr/options/) from the author of `Knitr` and `Rmarkdown` R packages.

Here is a short list of the most common that tend to be used by me.

```{r echo = F, message = F}
library(magrittr)

chunks <- readr::read_tsv("supplementary/chunk_options.tsv") %>%
       		dplyr::filter(this) %>%
       		dplyr::arrange(type) %>%
		      dplyr::select(-this)

chunks %>% knitr::kable()
```

--- There are way too many of them, good luck !

## Message and warning options

It is great that we have live code in our document, but results seems to be popluted with additional text. Those are "messages" from the package loading and some could be additional warning about casting values to different type. It depends, but generally suppressing message really helps with document apperance. I also tend to suppresss warnings, but this can cause issue in downstread analysis, be sure you understand your warnings before suppressing them.

Let's add a couple of options to our code chunk and press `Knit`

````markdown
`r ''````{r get_data, message = FALSE, warning = FALSE}
````

--- Much better, yeah?

## More code chunks and data wragnling

Let's add a couple more chunks and wrangle our data a little more. 

_It is okay if you don't understand some of R code in this example. The main point of these R chunks is to illustrate what can be done_

````markdown
## Getting and wrangling the data

`r ''````{r get_data}
make_url <- function(package) {
  paste0('<a href="https://cran.r-project.org/web/packages/', package, '">', package, '</a>')
}

library(magrittr)
library(dplyr)

avail_packages <- available.packages(contriburl = contrib.url("https://mirror.aarnet.edu.au/pub/CRAN/")) %>% 
                    as.data.frame() %>% 
                    as_tibble() %>% 
                    rowwise() %>% 
                    mutate(Depend = length(unlist(strsplit(Imports, split = ",")))) %>% 
                    ungroup() %>% 
                    select(Package, Version, Depend, Imports, License) %>% 
                    mutate(Package = make_url(Package),
                           Depend = ifelse(is.na(Imports), -1, Depend)) 
`r ''````

## Displaying table

`r ''````{r make_table, message=FALSE, warning=FALSE}
library(DT)
avail_packages %>% datatable(escape = FALSE)
`r ''````

## Plot the data

`r ''````{r plot}
library(ggplot2)
avail_packages %>% ggplot(aes(Depend)) + geom_bar()
`r ''````

````

--- Amazing !

## Session info

Let's include session info

```{r eval = F}
sessionInfo()
```

## Setup chunk

While it is great to be able to configure each chunk individually, but sometimes setting up a "global" chunk can help with making document chunks more manageable and easier to read. 

Let's include a "setup" chunk (name of the chunk can be anything) and set some of the options globally for the entire document. Note that we can also move all of our libraries into that chunk. This may or may not be a good idea, once again all depends what you need to show the person you are communicate this too. Don't forget to press `Knit`

Note that `include` option here set to `FALSE` meaning that chunk won't be included at all into the final document, as oppose to `eval = FALSE` where chunk is included but not evaluated.

````markdown
`r ''````{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(magrittr)
library(dplyr)
library(DT)
library(ggplot2)
`r ''````
````

--- How about that?

## Final look 

Just in case you've got lost, this is how our second section should look like

````markdown
# R code

`r ''````{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(magrittr)
library(dplyr)
library(DT)
library(ggplot2)
`r ''````

## Getting and wrangling the data

`r ''````{r get_data}

make_url <- function(package) {
  paste0('<a href="https://cran.r-project.org/web/packages/', package, '">', package, '</a>')
}

library(magrittr)
library(dplyr)

avail_packages <- available.packages(contriburl = contrib.url("https://mirror.aarnet.edu.au/pub/CRAN/")) %>% 
                    as.data.frame() %>% 
                    as_tibble() %>% 
                    rowwise() %>% 
                    mutate(Depend = length(unlist(strsplit(Imports, split = ",")))) %>% 
                    ungroup() %>% 
                    select(Package, Version, Depend, Imports, License) %>% 
                    mutate(Package = make_url(Package),
                           Depend = ifelse(is.na(Imports), -1, Depend)) 
`r ''````

## Displaying table

`r ''````{r make_table, message=FALSE, warning=FALSE}
library(DT)
avail_packages %>% datatable(escape = FALSE)
`r ''````

## Plot the data

`r ''````{r plot}
library(ggplot2)
avail_packages %>% ggplot(aes(Depend)) + geom_bar()
`r ''````
````
