# More Rmarkdown

We are going to inrease the difficulty a little bit and we are going to start working towards our [final documnet](examples/single_page_report.html).

Typically you will have some data set that you are trying to analyse and later present. There are likely to be some other prior steps before you get your tabular data. Those prior steps should also be documented. In this workshop we are going to start with a tabular data set straight away. We are going use [Domestic Airlines - On Time Performance](https://data.gov.au/data/dataset/domestic-airline-on-time-performance) data set from [data.gov.au](https://data.gov.au)

## Setup

First thing first is we need to download data. `read_csv()` function from [`readr`](https://readr.tidyverse.org/) package can "read" directly from url, but we are going to "cache" a file first and then we are going to reference our local copy. This will shorten our final html building time.
We should also check licence on the data set, especially if you are going to publish some of your analysis. This data is _Creative Commons Attribution 3.0 Australia_  licence, there is no problem in downloading and using the data.

Let's open new Rmarkdown file and delete everything from it except the yaml header.

### Setting global chunk options

As you have learned already you can manipulate each R chunk with options, but you can also set global settings for each chunk. Let's set `echo = TRUE` and `message = FALSE` globally. This means every R chunk will be echoed i.e shown in the final document and no messages will appear anywhere in the document.

```{r setup}
options(encoding="utf-8")

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

### Loading libraries

We are going to do our analysis with the help of [`tidyverse`](https://www.tidyverse.org/) library, let's load it in.

```{r eval = F}
library(tidyverse)
```

### Downloading the data

We are doing conditional download, so that we don't need re-download every time we build a document.

```{r}
fn_data <- "domestic_airline_performance.csv"
fn_notes <- "domestic_airline_performance_notes.txt"

if(!file.exists(fn_data)) {
  url_data <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/cf663ed1-0c5e-497f-aea9-e74bfda9cf44/download/otp_time_series_web.csv"
  url_notes <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/69e214b9-b66f-4636-9c84-3efda2d0d8ae/download/otptimeserieswebnotes.txt"

  download.file(url_data, fn_data)
  download.file(url_notes, fn_notes)
}

df <- read_csv(fn_data, quote = "")
df
```

## Exploring the data

Let's explore our data set. It is always helpful to get more information about the data set, good to start with getting dimensions of the `data.frame`, i.e number of rows and columns. Let's include the following bit of come into our Rmarkdown document. Note that instead of executing and showing the code we are stroing results in a variable `d`, and for that we are using `echo = FALSE` chunk option.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
	  "d <- df %>% dim",
	  "```")

cat(text, sep = '\n')
```

Later in the text we will be able to access variable as you would in R

```{r echo = F, comment=""}
text <- c("`r d[1]`")
cat(text, sep = '\n')
```

Lets add the following sentence to our Rmarkdown document and then `knit` to see the results.

```{r echo = F, comment=""}
text <- c("total number of observation `r d[1]` and total number of variables `r d[2]`")
cat(text, sep = '\n')
```

Now we are going to find out names of all airlines presented in the data set and number of observation for each airline. In this case we are not interested in showing actual code, so we are going to hide it with `echo = FALSE` once again.

<!--

Let's summarise our data to see how many observation each airline has.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
          "df %>%",
          "  select(Airline) %>%",
          "  distinct() %>%",
          "  arrange(Airline)",
	  "```")

cat(text, sep = '\n')
```
-->

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
          "df %>%",
          "  group_by(Airline) %>%",
          "  summarise(n = n()) %>%",
          "  arrange(-n)",
	   "```")

cat(text, sep = '\n')
```

I hope you have noticed "All Airlines" name in the `Airline` column. I am not fairly with such airlines. This is likely some summary field that we don't know about. In the real life you should consult people who has generated the data set, but in our case we are simply going to filter those observations out. Once again we are hiding the results from the final document, by setting `eval = TRUE` to make sure that the code runs since we will need `df2` later in the document.

```{r echo = F, comment=""}
text <- c("```{r echo = F, eval = T}",
	  "df2 <- df %>% filter(Airline != 'All Airlines')",
	  "```")

cat(text, sep = '\n')
```
## Including external files

- `child` chunk option allows us to include external files into our document

You have probably noticed that we have downloaded two files, the data set and the notes about the dat set. I didn't want to spend time going over those notes, but let's include them into our Rmarkdown document for future references.

Let's add the following text to out document

```{r echo = F, comment=""}
text <- c("### Additional information about the data set",
	  "",
	  "```{r child = 'domestic_airline_performance_notes.txt'}",
	  "",
	  "```",
	  "***")
cat(text, sep = '\n')
```

## Challenge: More Rmarkdown 1 {.exercise}

> 5 minutes

<details>
  <summary>
    1. Can you summarise routes in similar way as we did with airlines?
  </summary>
   use `group_by(Route)`
</details>

<details>
  <summary>
    2. Can spot an odd route in you summary? If you can filter is our from `df2`.
  </summary>
   `filter(Route != "All Ports-All Ports")`
</details>

## Visualising the data

Our document looks pretty good so far, let's add some visualisation to our document, with that will be covering these chunk options the following chunk options, all to do with figures manipulation.


- `fig.align` - left, right, center or default (left)
- `fig.height` - height specified in inches
- `fig.width` - width specified in inches
- `fig.cap` - string of text in quotes

Firstly lets make sure we have our data properly filtered. We are going to filter out `All Ports-All ports` routes, since this is similar to `All airlines` field and likely to be some summary field that we are not interested in. We also going to only look at two airlines, Jetstar and Qantas, due to time constrains. Here we don't need to set any particular chunk options because I think it would be informative to show our filtering and we actually need that code to be run.

```{r echo = F, comment=""}
text <- c("```{r}",
          "df2 <- df %>% filter(Airline == 'Jetstar' | Airline == 'Qantas')",
	  "                     Route != 'All Ports-All Ports')",
	  "```")

cat(text, sep = '\n')
```

Here we are summarising our data so that we have an idea of how many times a particular location had be used by our airlines per year. Let's include that bit code into our Rmarkdown document and once again let's hide our code from the document by setting `echo = FALSE`

```{r echo = F, comment=""}
text <- c("```{r echo = FALSE}",
          "df2 %>%",
          "  group_by(Airline, Year, Departing_Port) %>%",
          "  summarise(n = n()) %>%",
          "  ungroup %>%",
          "    ggplot(aes(Departing_Port, n, color = Airline)) +",
          "    geom_boxplot() +",
          "    theme(axis.text.x=element_text(angle=45, hjust=1))",
          "```")

cat(text, sep = '\n')
```

Let's experiment with setting different dimensions to our figure. Lets set both height and width to 4. Remember that units for height and width are in inches.

```{r echo = F, comment=""}
text <- c("```{r echo = FALSE, fig.height = 4, fig.width = 4}",
          "df2 %>%",
          "  group_by(Airline, Year, Departing_Port) %>%",
          "  summarise(n = n()) %>%",
          "  ungroup %>%",
          "    ggplot(aes(Departing_Port, n, color = Airline)) +",
          "    geom_boxplot() +",
          "    theme(axis.text.x=element_text(angle=45, hjust=1))",
          "```")

cat(text, sep = '\n')
```

Now let's try to align our figure to the center by setting `fig.align = 'center'` and re-build our document.

Let's experiment with setting wight to 15 and height to 9. At some point though we are going to start hitting the physical limit of the html page.

## Challenge: More Rmarkdown 2 {.exercise}

> 5 minutes

<details>
  <summary>
    1. Can you align figure to the right??
  </summary>
  `fig.align = 'right'`
</details>

<details>
  <summary>
    2. Align figure to the center and add figure legend using `fit.cap` chunk option
  </summary>
  `fig.align = 'center', fig.cap = 'Figure 1: blah'`
</details>

<details>
  <summary>
    3. Can you add some emphasis to figure legend, e.g make important parts bold or underlined?
  </summary>
  Remember that figure legend is just a string of text and any text can be marked
</details>

## More plots

Let's attempt to see what is the distribution of cancellation in any given year

Note the `warning` message that comes up in the text, let's assume we understand it and let's just turn it off by setting `warning = F`


```{r echo = F, comment=""}
text <- c("```{r, fig.width = 14, fig.height = 9}",
          "df2 %>%",
          "  filter(Airline == 'Jetstar' | Airline == 'Qantas') %>%",
          "  select(Airline, Departing_Port, Cancellations, Year) %>%",
          "    ggplot(aes(Departing_Port, Cancellations, color = factor(Year))) +",
          "      geom_boxplot() +",
          "      facet_wrap(~Airline) +",
          "      theme(axis.text.x=element_text(angle= 45, hjust=1))",
          "```")

cat(text, sep = '\n')
```

## More chunk options

I'm going to share a couple of more code chunks, these are mainly cosmetic, some of you may never use them.

- `prompt=FALSE` i.e mimic _consosle_
- `comment=` remove hash symbol at the front of the output

For this example I'm going to use simple `for` loop. We are going to use this variable `sentence <- c("Let", "the", "computer", "do", "the", "work")`

```{r echo=FALSE, comment=""}
text <- c("```{r}",
          "sentence <- c('Let', 'the', 'computer', 'do', 'the', 'work')",
          "",
          "for(word in sentence){",
          "  print(word)",
          "}",
          "```")

cat(text, sep = "\n")
```

```{r}
sentence <- c("Let", "the", "computer", "do", "the", "work")

for(word in sentence){
  print(word)
}
```

Let's add `prompt=TRUE`


```{r prompt=TRUE}
sentence <- c("Let", "the", "computer", "do", "the", "work")

for(word in sentence){
  print(word)
}
```
