# More Rmarkdown

We are going to inrease the difficulty a little now. We are going to start working towards our [final documnet](example.html).

Typically you will have some data set that you are trying to analyse. There are likely to be some other prior steps before you get your tabular data. Those prior steps should also be documented. In this workshop we are going to start with a tabular data set. I found this rather interesting data set at [data.gov.au](https://data.gov.au), [Domestic Airlines - On Time Performance](https://data.gov.au/data/dataset/domestic-airline-on-time-performance) and I decided to investigate it a bit closer.

## Setup

First thing first is we need to download it. Note that `read_csv` from [`readr`](https://readr.tidyverse.org/) package can "read" directly from url, but I wasn't sure if everytime I compile Rmarkdown to html it would re-download the file or use chached version. I decide to store a local copy on my computer for speed and simplisity. You should always check a licence on the data set, especially if you are going to publish some of your analysis. This data is _Creative Commons Attribution 3.0 Australia_  licence, there is no problem in downloading and using the data.


Let's open new Rmarkdown file and delete everything from it except the yaml header.

### Setting global chunk options

As you have learned already you can manipulate each R chunk with options, but you can also set global settings for each chunk. Let's set `echo = TRUE` and `message = FALSE` globally. This means every R chunk will be echoed i.e shown in the finale document and no messages will appear anywhere in the document.

```{r setup}
options(encoding="utf-8")

knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

### Loading libraries

We are going to do our analysis with the help of [`tidyverse`](https://www.tidyverse.org/) library, that in itself includes many other libraries.


```{r eval = F}
library(tidyverse)
```

### Downloading the data

We are doing conditional download here, don't re-download if we already have the file.

```{r}
fn_data <- "domestic_airline_performance.csv"
fn_notes <- "domestic_airline_performance_notes.txt"

if(!file.exists(fn_data)) {
  url_data <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/cf663ed1-0c5e-497f-aea9-e74bfda9cf44/download/otp_time_series_web.csv"
  url_notes <- "https://data.gov.au/data/dataset/29128ebd-dbaa-4ff5-8b86-d9f30de56452/resource/69e214b9-b66f-4636-9c84-3efda2d0d8ae/download/otptimeserieswebnotes.txt"

  download.file(url_data, fn_data)
  download.file(url_notes, fn_notes)
}

df <- read_csv(fn_data, quote = "")
df
```

## Challenge: 1 {.exercise}

> 2 minutes

<details>

  <summary>
    1. Exploring and familiarising yourself with the data set
  </summary>
  ```{r child="domestic_airline_performance_notes.txt"}

  ```
</details>


## Exploring the data

Now that we've got the data lets explore it. It always helps if we can find more information about the data set, particular what information each column might have.

Let's understand a bit better our data set by firstly figuring out how many observations and different variables we have. Instead of executing and showing the code we instead going to keep the chunk "silent" but still executing the R code and we are going to reuse our `d` variable later in the document.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
	  "d <- df %>% dim",
	  "```")

cat(text, sep = '\n')
```

Later in the text we'll access our `d` variable like you would in R

```{r echo = F, comment=""}
text <- c("`r d[1]`")
cat(text, sep = '\n')
```

Lets add the following sentence to our Rmarkdown document and then `knit` to see the results.

```{r echo = F, comment=""}
text <- c("total number of observation `r d[1]` and total number of variables `r d[2]`")
cat(text, sep = '\n')
```

Now we are going to look at total number of airlines, include the following code into your Rmarkdown document.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
          "df %>%",
           "select(Airline) %>%",
           "distinct() %>%",
           "arrange(Airline)",
	   "```")

cat(text, sep = '\n')
```

Let's summarise our data to see how many observation each airline has.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
          "df %>%",
           "group_by(Airline) %>%",
            "summarise(n = n()) %>%",
            "arrange(-n)",
	   "```")

cat(text, sep = '\n')
```

I hope you have noticed "All Airlines" name in the `Airline` column. I'm fairly certain that this isn't any specif airlines. In general we would need to consult people who we got the data from, but in our case we are simply going to filter those observations out.

```{r echo = F, comment=""}
text <- c("```{r echo = F}",
	  "df2 <- df %>% filter(Airline != 'All Airlines'",
	  "```")

cat(text, sep = '\n')
```


## Challenge: 1 {.exercise}

> 5 minutes

<details>
  <summary>
    1. Can you summarise routes in similar way as we did with airlines?
  </summary>
   use `group_by(Route)`
</details>

<details>
  <summary>
    2. Can spot an odd route in you summary? If you can filter is our from `df2`.
  </summary>
   `filter(Route != "All Ports-All Ports")`
</details>

## Visualising the data

In this section we are going to learn a few more chunk options, all to do with figure manipulation.

- `fig.align`
- `fig.cap`
- `fig.height`
- `fig.width`

Firstly lets make sure we have our data properly filtered, just in case you missed the challenge above

```{r echo = F, comment=""}
text <- c("```{r}",
          "df2 <- df %>% filter(Airline != 'All Airlines', Route != 'All Ports-All Ports')",
	  "```")

cat(text, sep = '\n')
```

Here we are summarising so that we have an idea of how many times a particular location had be use per airline per year and we are only going to look at two airlines, Jetstar and Qantas.

```{r echo = F, comment=""}
text <- c("```{r}",
          "df2 %>%",
          "  group_by(Airline, Year, Departing_Port) %>%",
          "  summarise(n = n()) %>%",
          "  ungroup %>%",
          "  filter(Airline == 'Jetstar' | Airline == 'Qantas') %>%",
          "    ggplot(aes(Departing_Port, n, color = Airline)) +",
          "    geom_boxplot() +",
          "    theme(axis.text.x=element_text(angle=45, hjust=1))",
          "```")

cat(text, sep = '\n')
```

In any given year what is the distribution of cancellation

Note the `warning` message that comes up in the text, let's assume we understand it and let's just turn it off by setting `warning = F`


```{r echo = F, comment=""}
text <- c("```{r, fig.width = 14, fig.height = 9}",
          "df2 %>%",
          "  filter(Airline == 'Jetstar' | Airline == 'Qantas') %>%",
          "  select(Airline, Departing_Port, Cancellations, Year) %>%",
          "    ggplot(aes(Departing_Port, Cancellations, color = factor(Year))) +",
          "      geom_boxplot() +",
          "      facet_wrap(~Airline) +",
          "      theme(axis.text.x=element_text(angle= 45, hjust=1))",
          "```")

cat(text, sep = '\n')
```
