# The R chunk 2

## Working with code chunks

Let's continue our exploration of chunk options and now try a different example using `mtcars` dataset and learn a few more chunk options.

```{r echo=FALSE, comment=""}
text1 <- c("```{r}",
           "summary(mtcars)",
           "```")

cat(text1, sep = "\n")
```

```{r}
summary(mtcars)
```

**Remember** You can go between Rmarkdown and _console_, to check your code, at any time. You should see your code block is highlighted differently and you should see a green arrow at the right hand site of that block. Press the green arrow to get an output in the _console_. You can also use `ctrl+enter` to do the same with the keyboard short cut.


```{r echo=FALSE, comment=""}
text1 <- c("```{r}",
           "plot(mtcars)",
           "```")

cat(text1, sep = "\n")
```

```{r}
plot(mtcars)
```

This is great, but a bit too much information, lets just focus on number of cylinders and hourse power.


```{r echo=FALSE, comment=""}
text1 <- c("```{r}",
           "plot(mtcars$cyl, mtcars$hp)",
           "```")

cat(text1, sep = "\n")
```

```{r}
plot(mtcars$cyl, mtcars$hp)
```

We can add a bit more information to our plot, to make more self descriptive.

```{r echo=FALSE, comment=""}
text1 <- c("```{r}",
           "plot(mtcars$cyl,
           mtcars$hp,
           main='House power vs number of cylinders',
           xlab = 'Numer of cylinders',
           ylab='House power')",
           "```")

cat(text1, sep = "\n")
```

```{r}
plot(mtcars$cyl,
     mtcars$hp,
     main='House power vs number of cylinders',
     xlab = 'Numer of cylinders',
     ylab='House power')
```

Here is a good example where we can hide our code from the viewer, since it isn't most interesting bit about this data. Let's turn `echo=FALSE` options for all our plots below.

Properly labelled plots are very informative, let's do that as well, starting with a title `main="Travelling speed vs Breaking distance"` and then labelling axis, x `xlab="Travelling speed (mhp)"` and y `ylab="Stopping distance (ft)"`


```{r, echo = FALSE}
plot(mtcars$cyl,
     mtcars$hp,
     main='House power vs number of cylinders',
     xlab = 'Numer of cylinders',
     ylab='House power')
```

We are no longer seeing the code, rather just the figure. You can try `eval = FALSE` by yourself to see what happens.

## Figures related chunk options

- `fig.align` - left, right, center or default (left)
- `fig.height` - height specified in inches
- `fig.width` - width specified in inches
- `fig.cap` - string of text in quotes

Let me show you a how to resize the plot with `fig.height` and `fig.width` and then we are going to do a challenge.

```{r echo=FALSE, comment=""}
text1 <- c("```{r fig.height = 4, fig.width = 4}",
           "plot(mtcars$cyl,
           mtcars$hp,
           main='House power vs number of cylinders',
           xlab = 'Numer of cylinders',
           ylab='House power')",
           "```")

cat(text1, sep = "\n")
```

```{r fig.height = 4, fig.width = 4}
plot(mtcars$cyl,
     mtcars$hp,
     main='House power vs number of cylinders',
     xlab = 'Numer of cylinders',
     ylab='House power')
```

Let's try to make it very big, by trying to set height and width to 15 inches.

```{r fig.height = 15, fig.width = 15}
plot(mtcars$cyl,
     mtcars$hp,
     main='House power vs number of cylinders',
     xlab = 'Numer of cylinders',
     ylab='House power')
```

Note that we are starting to hit "boundaries" of the documents. (want to talk about the fact that plot sits inside a `<div>` box)


## Challenge: code chunks {.exercise}

> 5 minutes

<details>
  <summary>
    1. Can you align figure to the right??
  </summary>
  `fig.align = 'right'`
</details>

<details>
  <summary>
    2. Align figure to the center and add figure legend
  </summary>
  `fig.align = 'center', fig.cap = 'Figure 1: blah'`
</details>

<details>
  <summary>
    3. Can you add some emphasis to figure legend, e.g make important parts bold or underlined?
  </summary>
  Remember that figure legend is just a string of text and any text can be marked
</details>

## More useful chunk options

I'm going to share a few more useful code chunks, some are cosmetic, some you may never use, but hey can be handy in making your document visually different.

- `prompt=FALSE` i.e mimic _consosle_
- `comment=` remove hash symbol at the front of the output
- `child=` path to another `Rmd` file
- `warning=FALSE`
- `messages=FALSE`

For this example I'm going to use simple `for` loop. We are going to use this variable `sentence <- c("Let", "the", "computer", "do", "the", "work")`

```{r echo=FALSE, comment=""}
text <- c("```{r}",
          "sentence <- c('Let', 'the', 'computer', 'do', 'the', 'work')",
          "",
          "for(word in sentence){",
          "  print(word)",
          "}",
          "```")

cat(text, sep = "\n")
```

```{r}
sentence <- c("Let", "the", "computer", "do", "the", "work")

for(word in sentence){
  print(word)
}
```

Let's add `prompt=TRUE`


```{r prompt=TRUE}
sentence <- c("Let", "the", "computer", "do", "the", "work")

for(word in sentence){
  print(word)
}
```

Now let's add an external Rmd content into this file using `child` option


```{r echo=FALSE, comment=""}
text <- c("```{r child = 'child_chunk_example.Rmd'}",
          "sentence <- c('Let', 'the', 'computer', 'do', 'the', 'work')",
          "",
          "for(word in sentence){",
          "  print(word)",
          "}",
          "```")

cat(text, sep = "\n")
```

```{r child = 'child_chunk_example.Rmd'}

sentence <- c("Let", "the", "computer", "do", "the", "work")

for(word in sentence){
  print(word)
}
```

## Challenge: more code chunks {.exercise}

> 5 minutes

<details>
  <summary>
    1. Add `library(tidyverse)` to get lots of messages and try to suppress them with chunk options
  </summary>
  `message = FALSE`
</details>

<details>
  <summary>
    2. In the example about `child_chunk_example.R` gives warning messages, can you suppress them from the output
  </summary>
  `message = FALSE`

## References

- [Section 2.6](https://bookdown.org/yihui/rmarkdown/r-code.html)
- [Pimp my RMD](https://holtzy.github.io/Pimp-my-rmd/)
