
## Where do you get bibtex file?

There are two ways to get `.bib` files:

- manually curate one: remember that this is a plain text file format, so you can just type it out in a text editor and save as `.bib` file
- export out of your citation management or the publication website

I tried a couple of website, [pubmed](https://www.ncbi.nlm.nih.gov/pubmed/) and [nature.com](https://www.nature.com), neither of those appear to allow you to export `.bib`. Pubmed exports `.nbib`, which is a different file format if you look inside. Nature exports `.ris`, again not the one that we want. [Google scholar](https://scholar.google.com.au/) has a button **import into Bibtext** that will give the correct citation.

As I've mentioned above all of those file formats are [actually supported](https://github.com/jgm/pandoc-citeproc/blob/master/man/pandoc-citeproc.1.md) the only problem is figuring out which "key" you can use to include your citation into the document. And there isn't much resource out there to help you with other bibliography file formats. For now I would say the best and easiest option is to stick with `.bib` format.

Let's practice including citation into our Rmarkdown document. This is the paper that we are going to cite _Excuse me, do you have a moment to talk about version control?_. Go to this URL [https://peerj.com/preprints/3159/](https://peerj.com/preprints/3159/), look for "Download" button on the right hand site, from the drop down select "BibTex". This will download citation file. On my computer the file name was `peerj-preprints-3159.bib`.

If you are curious to take a pick at the file content on your local computer, you can open such file with a text editor. Do not attempt to open with any other programs, such ash MS word or citation management tools. Most likely your computer will want to do just that, do not succumb! Mac people can use "TextMate" and windows people can use "Notepad" text editors. But this isn't necessary.

One you have `.bib` file, let's upload it into `rstudio.cloud`. Navigate to your "Files" panel, there you should see "Upload" button. Use that menu to upload your `.bib` file onto `rstudio.cloud`. You should see `.bib` file in your file browser if you are successful.
If you double click on that file it should open up in a different panel, text editor panel, and you should see the content of that file as shown below.

```{r echo = F, comment = ""}
text <- readr::read_file("supplementary/peerj-preprints-3159.bib")
cat(text, sep = "")
```

I hope every can identify "key" for this citation - `10.7287/peerj.preprints.3159v2`. To me this is very annoying key to use since it is not very memorable and hard to type. Let's edit this file in place and change that key to `Bryan2017`. Remember that key has to be unique per `.bib` file, but they can be any text. Make sure you save your changes (`ctrl^S`) and you can close this file now.

Let's include a reference to this `.bib` file in our Rmarkdown document via yaml header, using `bibliography` key.

```
---
bibliography: "peerj-preprints-3159.bib"
---
```

Now let's use our article key to cite in the text `[@Bryan2017]` and re-compile our html document.

## Challenge: Bibliographies 2 {.exercise}

> 5 minutes

<details>

  <summary>
    1. Go to [google scholar](https://scholar.google.com.au/) and search for that same article _Excuse me, do you have a moment to talk about version control?_, find and export `.bib` citation and include it into our Rmarkdown document.
  </summary>
  Relatively hard task since I haven't explained how to include multiple `.bib` files into yaml header.

</details>

<details>

  <summary>
    2. If you are manually curating `.bib` file, how do you know which types and fields are allowed?
  </summary>
   This resource can be [useful](#bibtex)

</details>

## More about `.bib` and Rmarkdown

As was mentioned above one can append multiple references into a single `.bib` file or you can provide a list of `.bib` file

```
---
bibliography: ["peerj-preprints-3159.bib", "another_reference.bib"]
---
```

OR

```
---
bibliography:
  - "peerj-preprints-3159.bib"
  - "another_reference.bib"
---
```

If you want to reference multiple articles in the text each key should be separated by a semicolon (`;`)


```
[@Bryan2017; @bryan2018excuse]
```

You can include any text inside square brackets

```
[see @Bryan2017 p 12; also this ref @bryan2018excuse]
```

If you would like to suppress author's name you can use minus sign


```
[-@Bryan2017; -@bryan2018excuse]
```

You can also enable hyperlinking of the citation to the corresponding entry in the references as follows

```
---
link-citations: true
---
```

## Citing R packages

R provides convenient `citation()` function that one can use to cite R packages e.g

```
citation("ggplot2")
```

## Challenge: Bibliographies 3 {.exercise}

> 2 minutes

<details>

  <summary>
    1. Can you get `.bib` citation for `bookdown` package?
  </summary>
  You don't get `.bib` text content right away, but if you are careful at reading the message you will understand that you need to do this

   ```{r eval = F}
   citation("bookdown") %>% print(bibtex = TRUE)
   ```

</details>
