# Bookdown

In this section we are going just touch on `bookdown::bookdown_site`, which is a custom "function" for generating multi-paged document, books, thesis and other documents.

## Introduction

Thus far in this part of the book we have been using default `site` generation function

```YAML
site: rmarkdown::default_site
```

Bookdown package provide another convenient function to build multi-paged book or alike documents by setting `site` key to the following value in our YAML `_site.yml` file. Let's add the following code into our `_site.ytml` file and press "Build"

```YAML
site: bookdown::bookdown_site
```

## Masking files

As was mention previous, any files with an `_` at the front won't be included into our site. Since presentations are well suited for the book format, let's suppress `presenting.Rmd` file using the following command. Alternative use point and click navigation in the "File" tab to rename the file.

```{r eval = F}
file.rename("presenting.Rmd", "_presenting.Rmd")
```

## Adding section headers

Let's add a couple of section headers into `index.Rmd` and `about.Rmd` files and press "Build" again.

## Bookdown skeleton

### Apporach 1

This approach wont work on the desktop (and server) clients. 

```
File
 └── New Project
 
Then select "New Directory" -> "Book project using bookdown"
```

Everything will be setup, including building tools. Once the project loaded press "Build" button

### Approach 2

Since rstudio.cloud automatically starts as a RStudio project, we are not able to select which project "type" we would like. However we can exectu the following command, which will give the same overall results.

```{r eval = F}
bookdown:::bookdown_skeleton(".")
```
